### 2022-06-14 chapters/interpretability/imli/baseline.tex
minor edits
preliminaries notations updated

\subsection{Description of Variables}
\label{interpretability_imli_sec:variables} 
We initially preprocess $ \mathbf{X} $  to account for the negation of  Boolean features while learning a classifier.\footnote{This preprocessing is similarly applied in  \cite{malioutov2013exact}.} In the preprocessing step, we negate each column in $ \mathbf{X} $ to a new column and append it to $ \mathbf{X} $. For example, if ``$ \text{age }\ge 25 $'' is a Boolean feature, we add another feature ``$ \text{age }< 25 $'' in  $ \mathbf{X} $ by negating the column ``$ \text{age }\ge 25 $''. Hence, in the rest of the chapter, we refer $ m $ as the modified number of columns in $ \mathbf{X} $. We next discuss the variables in the MaxSAT problem.

We consider two types of Boolean variables: (i) \emph{feature} variables $ b $ corresponding to input features and (ii) \emph{error} variables $ \eta $ corresponding to the classification error of samples. We define a Boolean variable $ b^i_j $ that becomes true if feature $ x_j $ appears in the $ i^\text{th} $ clause of $ \Rule $, thereby contributing to an increase in the rule-size of $ \Rule $, and $ b^i_j $ is assigned false, otherwise. Moreover, we define an error variable $ \eta_l $ to attribute to whether the sample $ \mathbf{X}_l $ is classified correctly or not. Specifically, $ \eta_l $ becomes true if $ \mathbf{X}_l $ is misclassified, and becomes false otherwise.  We next discuss the MaxSAT encoding to solve the classification problem.

\subsection{MaxSAT Encoding}
\label{interpretability_imli_sec:encoding}
We consider a partial-weighted MaxSAT formula, where we encode the objective function in Eq.~\eqref{interpretability_imli_eq:obj} as soft clauses. For each sample in $ \mathbf{X} $, we construct hard clauses specifying that the sample is either  classified correctly and $ \eta_l $ is false or it is misclassified and $ \eta_l $ is true. We next discuss the MaxSAT encoding in detail.

@@

	\item \textbf{Soft clauses for minimizing rule-sparsity:} To favor rule-sparsity, {\imli} tries to learn a classifier with as few literals as possible. Hence, for each feature variable $ b_j^i $, {\imli} constructs a unit clause as $ \neg b_j^i $. Similar to training accuracy, the weight for this clause is derived as $ \lambda $ from Eq.~\eqref{interpretability_imli_eq:obj}.
	\begin{equation}
		\label{interpretability_imli_soft_clause_feature}
		S_j^i:= \neg b_j^i; \qquad \mathsf{wt}(S_j^i)= \lambda
	\end{equation}

	\item \textbf{Hard clauses for encoding constraints:} In a MaxSAT problem, constraints that must be satisfied are encoded as hard clauses. In {\imli}, we have a constraint that, if the error variable is false, the associated sample must be correctly classified, and vice-versa. Let  $\mathbf{b}_i= \{b^i_{j} \mid j \in \{1,\dots,m \}\}$ be a vector of feature variables corresponding to the $ i^\text{th} $  clause in $ \mathcal{R} $. Then, we define the following hard clause.

	\begin{equation}
			\label{interpretability_imli_hard_clause_rule_consistency}
		H_l:= \neg \eta_l \rightarrow \Big(y_l\leftrightarrow \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}\Big); \qquad  \mathsf{wt}(H_l)=\infty
	\end{equation}

	In the hard clause, $ \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}\ $ is a CNF formula including variables $ b_j^i $ for which the associated  feature-value is $ 1 $ in  $ \mathbf{X}_l $. Since $ y_l \in \{0,1\} $ is a constant, the constraint to the right of the implication ``$ \rightarrow $'' is either  $ \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}\ $ or its complement. Therefore, the hard clause enforces that if the sample is correctly classified (using $ \neg \eta_l $), either  $ \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}\ $ or its complement is true depending on the class-label of the sample.  We highlight that according to the definition of ``$ \rightarrow $'', the left constraint $ \neg \eta_l $ can  be false while the right constraint of 	``$ \rightarrow $'' is true.  This, however, incurs unnecessarily dissatisfying the soft clause $ E_l $, which is a sub-optimal solution and hence this solution is not returned by the MaxSAT solver. Therefore, the single-implication ``$ \rightarrow $'' in the hard clause $ H_l $ acts as a double-implication ``$ \leftrightarrow $''  due to the soft clause $ E_l $.

@@

\paragraph{Translating $ E_l, S^i_j, H_l $ to a CNF formula.} 
The soft clauses $ E_l $ and $ S_j^i $ are unit clauses and hence, no translation is required for them.  In the hard clause, when $ y_l=1 $, the simplification is $ H_l:= \neg \eta_l \rightarrow \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}  $. In this case, we apply the equivalence rule in propositional logic $ (a \rightarrow b) \equiv (\neg a \vee b) $  to encode $ H_l $ into  CNF. In contrast, when $ y_i=0 $, we simplify the hard clause as $ H_l:= \neg \eta_l \rightarrow \neg (\bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}) \Rightarrow  \neg \eta_l \rightarrow \bigvee_{i=1}^k \neg({\mathbf{X}_l} \circ {\mathbf{b}_{i}})$. Since $ \mathbf{X}_l \circ \mathbf{b}_{i} $ constitutes a disjunction of literals, we apply Tseytin transformation to encode  $ \neg (\mathbf{X}_l \circ \mathbf{b}_{i}) $ into CNF. More specifically, we introduce an auxiliary variable $ z_{l,i} $ corresponding to the clause $ \neg ({\mathbf{X}_l} \circ {\mathbf{b}_{i}}) $. Formally, we replace $ H_l := \neg \eta_l \rightarrow  \bigvee_{i=1}^k \neg ({\mathbf{X}_l} \circ {\mathbf{b}_{i}}) $ with   $ \bigwedge_{i=0}^k H_{l,i} $,  where $ H_{l,0}:= (\neg \eta_l \rightarrow  \bigvee_{i=1}^k  z_{l,i}) $  and  $  H_{l,i}:= z_{l,i} \rightarrow  \neg({\mathbf{X}_l} \circ {\mathbf{b}_{i}})  $ for $ i=\{1,\dots,k\} $. Finally, we  apply the equivalence of $ (a \rightarrow b) \equiv (\neg a \vee b) $ on $ H_{l,i} $ to translate them into CNF. For either value of $ y_l $, the weight of each translated hard clause in the CNF formula is $ \infty $.

% Now,  the hard clause can be simplified as $ H_l:= \neg \eta_l \rightarrow \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}  $ when $ y_l=1 $, otherwise  $ H_l:= \neg \eta_l \rightarrow \neg (\bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}) \Rightarrow  \neg \eta_l \rightarrow \bigvee_{i=1}^k \neg({\mathbf{X}_l} \circ {\mathbf{b}_{i}})$.

% In the hard clause,  $\bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}  $ is already in CNF. When $ y_l=1 $, the constraint $ H_l $ can be directly encoded as CNF by using equivalence of $ (a \rightarrow b) \equiv (\neg a \vee b) $. When $ y_i=0 $, we apply

@@

Any off-the-shelf MaxSAT solver can output an optimal assignment $ \sigma^* $ of the MaxSAT query $ (Q, \weight{\cdot}) $. We extract $ \sigma^* $ to construct the classifier $ \Rule $ and compute training errors as follows.
\begin{construction}
	\label{interpretability_imli_construction:rule}
	Let $\sigma^* = \MaxSAT(Q,\weight{\cdot})$. Then $x_j \in {\clause(\Rule,i)}$ if and only if $\sigma^*(b_{j}^{i}) = 1$. Additionally, $ \mathbf{X}_l $ is misclassified if and only if $ \sigma^*(\eta_l) = 1 $.
\end{construction}

@@

\subsubsection*{Example of encoding}

We illustrate the MaxSAT encoding for a toy example consisting of four samples and two binary features. Our goal is to learn a two clause CNF classifier that can approximate the training data. 
\[ 
\mathbf{X}_\mathsf{orig} = 
\begin{blockarray}{cc}
x_1 & x_2 \\
\begin{block}{[cc]}
0 & 0 \\
0 & 1 \\
1 & 0 \\
1 & 1 \\
\end{block}
\end{blockarray} \implies 
\mathbf{X} = 
\begin{blockarray}{cccc}
x_1 & \neg x_1 & x_2  & \neg x_2\\
\begin{block}{[cccc]}
0 & 1 & 0 & 1\\
0 & 1 & 1 & 0\\
1 & 0 & 0 & 1\\
1 & 0 & 1 & 0\\
\end{block}
\end{blockarray}; \quad 
\mathbf{y} = 
\begin{bmatrix} 
1 \\ 0 \\ 0 \\ 1
\end{bmatrix}
\]
In the preprocessing step, we negate the columns in $ \mathbf{X}_\mathsf{orig} $ and add complemented features $ \{\neg x_1, \neg x_2\} $ in $ \mathbf{X} $. In the MaxSAT encoding, we define $ 8 $ feature variables ($ 4 $ features $ \times $ $ 2 $ clauses in the classifier) and $ 4 $ error variables. For example, for feature $ x_2 $, introduced feature variables are $ \{b^1_3, b^2_3\} $ and for feature $ \neg x_2 $, introduced variables are $ \{b^1_4, b^2_4\} $. For four samples, error variables are $ \{\eta_1, \eta_2, \eta_3, \eta_4\} $. We next show the soft and hard clauses in the MaxSAT encoding

\[
\begin{split}
&E_1:= (\neg \eta_1); \quad
E_2:= (\neg \eta_2); \quad E_3:= (\neg \eta_3); \quad
E_4:= (\neg \eta_4)\\
& S_1^1 := (\neg b_1^1);\quad 
 S^1_2 := (\neg b^1_2);\quad 
 S^1_3 := (\neg b^1_3);\quad
 S^1_4 := (\neg b^1_4);\quad \\ 
& S_1^2 := (\neg b_1^2);\quad 
S^2_2 := (\neg b^2_2);\quad 
S^2_3 := (\neg b^2_3);\quad 
S^2_4 := (\neg b^2_4);\quad\\ 
& H_1:= (\neg \eta_1 \rightarrow ((b_2^1 \vee b_4^1)\wedge (b_2^2 \vee b_4^2)));\\
& H_2:= (\neg \eta_2 \rightarrow (\neg(b_2^1 \vee b_3^1) \vee \neg(b_2^2 \vee b_3^2)));\\
& H_3:= (\neg \eta_3 \rightarrow (\neg (b_1^1 \vee b_4^1) \vee \neg(b_1^2 \vee b_4^2)));\\
& H_4:= (\neg \eta_4 \rightarrow ((b_1^1 \vee b_3^1) \wedge (b_1^2 \vee b_3^2)));\\
\end{split}
\]

In this example, we consider regularizer $ \lambda = 0.1 $, thereby setting the weight on accuracy as $ \weight{E_i} = 1 $ and rule-sparsity weight as $ \weight{S_j^i} = 0.1 $. Intuitively, the penalty for misclassifying a sample is $ 10 $ times more than the penalty for adding a feature in the classifier. For this MaxSAT query, the optimal solution classifies all samples correctly by assigning four feature variables $ \{b^1_1, b^1_4, b^2_2, b^2_3\} $ to true. Therefore, by applying Construction~\ref{interpretability_imli_construction:rule}, the classifier is $ (x_1 \vee \neg x_2)  \wedge (\neg x_1 \vee x_2) $.\footnote{The presented MaxSAT-based formulation does not learn a CNF classifier with both $ x_i $ and $ \neg x_i $ in the same clause. The reason is that a clause with both $ x_i $ and $ \neg x_i $ connected by OR $ (\vee) $ does not increase accuracy but increases rule-size and hence, this is not an optimal classifier.}

\subsection{Learning with Non-binary Features}
	\label{interpretability_imli_sec:non-binary}
	The presented MaxSAT encoding requires input samples to have binary features. Therefore, we discretize datasets with categorical and continuous features into binary features. For each continuous feature, we apply equal-width discretization that splits the feature into a fixed number of bins. For example, consider a continuous feature $ x_c \in [a,b] $. In discretization, we split the domain $ [a,b] $ into three bins with two split points $ \{a',b'\} $ such that $ a<a'<b'<b $.  Therefore, the resulting three discretized features are $ { a \le x_c < a'} $, $ {a' \le x_c < b'} $, and $ { b' \le x_c  \le b } $. An alternate to this close-interval discretization is open-interval discretization, where we consider six discretized features with each feature being compared with one threshold. In that case, the discretized features are: $ x_c \ge a, x_c \ge a', x_c \ge b', x_c < a', x_c < b', x_c \le b $. Both open-interval and close-interval discretization techniques have their use-cases where one or the other is appropriate. For simplicity, we experiment with close-interval discretization in this chapter.

@@

\subsection{Flexible Interpretability Objectives}
\label{interpretability_imli_sec:complex_interpretability_objectives}
{\imli} can be extended to consider more flexible interpretability objectives than the simplified one in Eq.~\eqref{interpretability_imli_eq:obj}. In Eq.~\eqref{interpretability_imli_eq:obj}, we prioritize all features equally by providing the same weight to the clause $ S_j^i $ for all  values of $ i $ and $ j $. In practice, users may prefer rules containing certain features. In {\imli}, such an extension can be achieved by modifying the weight function and/or the definition of  $ S_j^i $. For example, to constrain the classifier to never include a feature, the weight of the clause $ S_j^i := \neg b_j^i $ can be set to $ \infty $. In contrast, to always include a feature, we can define $ S_j^i := b_j^i $ with weight $ \infty $. In both cases, we treat $ S_j^i $ as a hard clause.

Another use case may be to learn rules where clauses have disjoint set of features. To this end, we consider a pseudo-Boolean constraint $ \sum_{i=1}^k b^i_j \le 1 $, which specifies that the feature $ x_j $ appears in at-most one of the $ k $ clauses. This constraint may be soft or hard depending on the priority in the application domain. In either case, we convert this constraint to CNF using pseudo-Boolean to CNF translation~\cite{philipp2015pblib}. Thus, {\imli} allows us to consider varied interpretability constraints by only modifying the MaxSAT query without requiring changes in the MaxSAT solving. Therefore, the \emph{separation between modeling and solving} turns out to be the key strength of {\imli}.

>>>>

\subsection{Description of Variables}
\label{interpretability_imli_sec:variables} 
We initially preprocess feature vector $ \mathbf{x} $  to account for the negation of  Boolean features while learning a classifier.\footnote{This preprocessing is similarly applied in  \cite{malioutov2013exact}.} In the preprocessing step, we negate each feature in $ \mathbf{x} $ to a new feature and append it to $ \mathbf{x} $. For example, if ``$ \text{age }\ge 25 $'' is a Boolean feature, we add another feature ``$ \text{age }< 25 $'' in  $ \mathbf{x} $ by negating the feature ``$ \text{age }\ge 25 $''. Hence, in the rest of the chapter, we refer $ m $ as the modified number of features in $ \mathbf{x} $. We next discuss the variables in the MaxSAT problem.

We consider two types of Boolean variables: (i) \emph{feature} variables $ \bool $ corresponding to input features and (ii) \emph{error} variables $ \eta $ corresponding to the classification error of samples. We define a Boolean variable $ \bool^i_j $ that becomes true if feature $ X_j $ appears in the $ i^\text{th} $ clause of $ \Rule $, thereby contributing to an increase in the rule-size of $ \Rule $, and $ \bool^i_j $ is assigned false, otherwise. Moreover, we define an error variable $ \eta_l $ to attribute to whether the $ l^\text{th} $ sample $ (\mathbf{x}^{(l)}, y^{(l)}) $ is classified correctly or not. Specifically, $ \eta_l $ becomes true if $ (\mathbf{x}^{(l)}, y^{(l)}) $ is misclassified, and becomes false otherwise.  We next discuss the MaxSAT encoding to solve the classification problem.

\subsection{MaxSAT Encoding}
\label{interpretability_imli_sec:encoding}
We consider a partial-weighted MaxSAT formula, where we encode the objective function in Eq.~\eqref{interpretability_imli_eq:obj} as soft clauses and the learning constraints as hard clauses. We next discuss the MaxSAT encoding in detail.

@@

	\item \textbf{Soft clauses for minimizing rule-sparsity:} To favor rule-sparsity, {\imli} tries to learn a classifier with as few literals as possible. Hence, for each feature variable $ \bool_j^i $, {\imli} constructs a unit clause as $ \neg \bool_j^i $. Similar to training accuracy, the weight for this clause is derived as $ \lambda $ from Eq.~\eqref{interpretability_imli_eq:obj}.
	\begin{equation}
		\label{interpretability_imli_soft_clause_feature}
		S_j^i:= \neg \bool_j^i; \qquad \mathsf{wt}(S_j^i)= \lambda
	\end{equation}

	\item \textbf{Hard clauses for encoding constraints:} In a MaxSAT problem, constraints that must be satisfied are encoded as hard clauses. In {\imli}, we have a constraint that, if the error variable is false, the associated sample must be correctly classified, and vice-versa. Let  $\boolset_i= \{\bool^i_{j} \mid j \in \{1,\dots,m \}\}$ be a vector of feature variables corresponding to the $ i^\text{th} $  clause in $ \mathcal{R} $. Then, we define the following hard clause.

	\begin{equation}
			\label{interpretability_imli_hard_clause_rule_consistency}
		H_l:= \neg \eta_l \rightarrow \Big(y^{(l)}\leftrightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\Big); \qquad  \mathsf{wt}(H_l)=\infty
	\end{equation}

	In the hard clause, $ \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\ $ is a CNF formula including variables $ \bool_j^i $ for which the associated  feature-value is $ 1 $ in  $ \mathbf{x}^{(l)} $. Since $ y^{(l)} \in \{0,1\} $ is a constant, the constraint to the right of the implication ``$ \rightarrow $'' is either  $ \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\ $ or its complement. Therefore, the hard clause enforces that if the sample is correctly classified (using $ \neg \eta_l $), either  $ \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\ $ or its complement is true depending on the class-label of the sample.  We highlight that according to the definition of ``$ \rightarrow $'', the left constraint $ \neg \eta_l $ can  be false while the right constraint of 	``$ \rightarrow $'' is true.  This, however, incurs unnecessarily dissatisfying the soft clause $ E_l $, which is a sub-optimal solution and hence this solution is not returned by the MaxSAT solver. Therefore, the single-implication ``$ \rightarrow $'' in the hard clause $ H_l $ acts as a double-implication ``$ \leftrightarrow $''  due to the soft clause $ E_l $.

@@

\paragraph{Translating $ E_l, S^i_j, H_l $ to a CNF formula.} 
The soft clauses $ E_l $ and $ S_j^i $ are unit clauses and hence, no translation is required for them.  In the hard clause, when $ y^{(l)}=1 $, the simplification is $ H_l:= \neg \eta_l \rightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}  $. In this case, we apply the equivalence rule in propositional logic $ (a \rightarrow b) \equiv (\neg a \vee b) $  to encode $ H_l $ into  CNF. In contrast, when $ y_i=0 $, we simplify the hard clause as $ H_l:= \neg \eta_l \rightarrow \neg (\bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}) \Rightarrow  \neg \eta_l \rightarrow \bigvee_{i=1}^k \neg({\mathbf{x}^{(l)}} \circ {\boolset_{i}})$. Since $ \mathbf{x}^{(l)} \circ \boolset_{i} $ constitutes a disjunction of literals, we apply Tseytin transformation to encode  $ \neg (\mathbf{x}^{(l)} \circ \boolset_{i}) $ into CNF. More specifically, we introduce an auxiliary variable $ z_{l,i} $ corresponding to the clause $ \neg ({\mathbf{x}^{(l)}} \circ {\boolset_{i}}) $. Formally, we replace $ H_l := \neg \eta_l \rightarrow  \bigvee_{i=1}^k \neg ({\mathbf{x}^{(l)}} \circ {\boolset_{i}}) $ with   $ \bigwedge_{i=0}^k H_{l,i} $,  where $ H_{l,0}:= (\neg \eta_l \rightarrow  \bigvee_{i=1}^k  z_{l,i}) $  and  $  H_{l,i}:= z_{l,i} \rightarrow  \neg({\mathbf{x}^{(l)}} \circ {\boolset_{i}})  $ for $ i=\{1,\dots,k\} $. Finally, we  apply the equivalence of $ (a \rightarrow b) \equiv (\neg a \vee b) $ on $ H_{l,i} $ to translate them into CNF. For either value of $ y^{(l)} $, the weight of each translated hard clause in the CNF formula is $ \infty $.

% Now,  the hard clause can be simplified as $ H_l:= \neg \eta_l \rightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}  $ when $ y^{(l)}=1 $, otherwise  $ H_l:= \neg \eta_l \rightarrow \neg (\bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}) \Rightarrow  \neg \eta_l \rightarrow \bigvee_{i=1}^k \neg({\mathbf{x}^{(l)}} \circ {\boolset_{i}})$.

% In the hard clause,  $\bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}  $ is already in CNF. When $ y^{(l)}=1 $, the constraint $ H_l $ can be directly encoded as CNF by using equivalence of $ (a \rightarrow b) \equiv (\neg a \vee b) $. When $ y_i=0 $, we apply

@@

Any off-the-shelf MaxSAT solver can output an optimal assignment $ \sigma^* $ of the MaxSAT query $ (Q, \clauseweight{\cdot}) $. We extract $ \sigma^* $ to construct the classifier $ \Rule $ and compute training errors as follows.
\begin{construction}
	\label{interpretability_imli_construction:rule}
	Let $\sigma^* = \MaxSAT(Q,\clauseweight{\cdot})$. Then $X_j \in {\clause(\Rule,i)}$ if and only if $\sigma^*(\bool_{j}^{i}) = 1$. Additionally, $ (\mathbf{x}^{(l)}, y^{(l)}) $ is misclassified if and only if $ \sigma^*(\eta_l) = 1 $.
\end{construction}

@@

\begin{example}{MaxSAT Encoding.}
\normalfont

We illustrate the MaxSAT encoding for a toy example consisting of four samples and two binary features. Our goal is to learn a two clause CNF classifier that can approximate the training data. 
\[ 
\dataset_\mathsf{orig} = 
\begin{blockarray}{ccc}
X_1 & X_2 & Y\\
\begin{block}{[ccc]}
0 & 0 & 1\\
0 & 1 & 0\\
1 & 0 & 0\\
1 & 1 & 1\\
\end{block}
\end{blockarray} \implies 
\dataset = 
\begin{blockarray}{ccccc}
X_1 & \neg X_1 & X_2  & \neg X_2 & Y\\
\begin{block}{[ccccc]}
0 & 1 & 0 & 1 & 1\\
0 & 1 & 1 & 0 & 0\\
1 & 0 & 0 & 1 & 0\\
1 & 0 & 1 & 0 & 1\\
\end{block}
\end{blockarray}
\]
In the preprocessing step, we negate the features in $ \dataset_\mathsf{orig} $ and add complemented features $ \{\neg X_1, \neg X_2\} $ in $ \dataset $. In the MaxSAT encoding, we define $ 8 $ feature variables ($ 4 $ features $ \times $ $ 2 $ clauses in the classifier) and $ 4 $ error variables. For example, for feature $ X_2 $, introduced feature variables are $ \{\bool^1_3, \bool^2_3\} $ and for feature $ \neg X_2 $, introduced variables are $ \{\bool^1_4, \bool^2_4\} $. For four samples, error variables are $ \{\eta_1, \eta_2, \eta_3, \eta_4\} $. We next show the soft and hard clauses in the MaxSAT encoding

\[
\begin{split}
&E_1:= (\neg \eta_1); \quad
E_2:= (\neg \eta_2); \quad E_3:= (\neg \eta_3); \quad
E_4:= (\neg \eta_4)\\
& S_1^1 := (\neg \bool_1^1);\quad 
 S^1_2 := (\neg \bool^1_2);\quad 
 S^1_3 := (\neg \bool^1_3);\quad
 S^1_4 := (\neg \bool^1_4);\quad \\ 
& S_1^2 := (\neg \bool_1^2);\quad 
S^2_2 := (\neg \bool^2_2);\quad 
S^2_3 := (\neg \bool^2_3);\quad 
S^2_4 := (\neg \bool^2_4);\quad\\ 
& H_1:= (\neg \eta_1 \rightarrow ((\bool_2^1 \vee \bool_4^1)\wedge (\bool_2^2 \vee \bool_4^2)));\\
& H_2:= (\neg \eta_2 \rightarrow (\neg(\bool_2^1 \vee \bool_3^1) \vee \neg(\bool_2^2 \vee \bool_3^2)));\\
& H_3:= (\neg \eta_3 \rightarrow (\neg (\bool_1^1 \vee \bool_4^1) \vee \neg(\bool_1^2 \vee \bool_4^2)));\\
& H_4:= (\neg \eta_4 \rightarrow ((\bool_1^1 \vee \bool_3^1) \wedge (\bool_1^2 \vee \bool_3^2)));\\
\end{split}
\]

In this example, we consider regularizer $ \lambda = 0.1 $, thereby setting the weight on accuracy as $ \clauseweight{E_i} = 1 $ and rule-sparsity weight as $ \clauseweight{S_j^i} = 0.1 $. Intuitively, the penalty for misclassifying a sample is $ 10 $ times more than the penalty for adding a feature in the classifier. For this MaxSAT query, the optimal solution classifies all samples correctly by assigning four feature variables $ \{\bool^1_1, \bool^1_4, \bool^2_2, \bool^2_3\} $ to true. Therefore, by applying Construction~\ref{interpretability_imli_construction:rule}, the classifier is $ (X_1 \vee \neg X_2)  \wedge (\neg X_1 \vee X_2) $.\footnote{The presented MaxSAT-based formulation does not learn a CNF classifier with both $ X_i $ and $ \neg X_i $ in the same clause. The reason is that a clause with both $ X_i $ and $ \neg X_i $ connected by OR $ (\vee) $ does not increase accuracy but increases rule-size and hence, this is not an optimal classifier.}

\end{example}

\subsection{Learning with Non-binary Features}
	\label{interpretability_imli_sec:non-binary}
	The presented MaxSAT encoding requires input samples to have binary features. Therefore, we discretize datasets with categorical and continuous features into binary features. For each continuous feature, we apply equal-width discretization that splits the feature into a fixed number of bins. For example, consider a continuous feature $ X_c \in [a,b] $. In discretization, we split the domain $ [a,b] $ into three bins with two split points $ \{a',b'\} $ such that $ a<a'<b'<b $.  Therefore, the resulting three discretized features are $ { a \le X_c < a'} $, $ {a' \le X_c < b'} $, and $ { b' \le X_c  \le b } $. An alternate to this close-interval discretization is open-interval discretization, where we consider six discretized features with each feature being compared with one threshold. In that case, the discretized features are: $ X_c \ge a, X_c \ge a', X_c \ge b', X_c < a', X_c < b', X_c \le b $. Both open-interval and close-interval discretization techniques have their use-cases where one or the other is appropriate. For simplicity, we experiment with close-interval discretization in this chapter.

@@

\subsection{Flexible Interpretability Objectives}
\label{interpretability_imli_sec:complex_interpretability_objectives}
{\imli} can be extended to consider more flexible interpretability objectives than the simplified one in Eq.~\eqref{interpretability_imli_eq:obj}. In Eq.~\eqref{interpretability_imli_eq:obj}, we prioritize all features equally by providing the same weight to the clause $ S_j^i $ for all  values of $ i $ and $ j $. In practice, users may prefer rules containing certain features. In {\imli}, such an extension can be achieved by modifying the weight function and/or the definition of  $ S_j^i $. For example, to constrain the classifier to never include a feature, the weight of the clause $ S_j^i := \neg \bool_j^i $ can be set to $ \infty $. In contrast, to always include a feature, we can define $ S_j^i := \bool_j^i $ with weight $ \infty $. In both cases, we treat $ S_j^i $ as a hard clause.

Another use case may be to learn rules where clauses have disjoint set of features. To this end, we consider a pseudo-Boolean constraint $ \sum_{i=1}^k \bool^i_j \le 1 $, which specifies that the feature $ X_j $ appears in at-most one of the $ k $ clauses. This constraint may be soft or hard depending on the priority in the application domain. In either case, we convert this constraint to CNF using pseudo-Boolean to CNF translation~\cite{philipp2015pblib}. Thus, {\imli} allows us to consider varied interpretability constraints by only modifying the MaxSAT query without requiring changes in the MaxSAT solving. Therefore, the \emph{separation between modeling and solving} turns out to be the key strength of {\imli}.

### 2022-06-14 chapters/interpretability/imli/problem_formulation.tex
preliminaries notations updated

 \section{Problem Formulation}
 \label{interpretability_imli_sec:problem}
Given (i) a binary feature matrix $ \mathbf{X} \in \{0,1\}^{n \times m} $ with $ n $ samples and $ m $ features, (ii) a class-label vector $ \mathbf{y} \in \{0,1\}^n $, (iii) a positive integer $ k \ge 1 $ denoting the number of clauses, and (iv) a regularization parameter $ \lambda \in \mathbb{R}^+ $, we learn a classifier $ \Rule $ represented as a $ k $-clause CNF formula that can separate samples of class $ 1 $ from class $ 0 $.

@@

 \[
 \min\limits_{\mathcal{R}} |\mathcal{R}|\text{ such that }\forall i, y_i=\mathcal{R}(\mathbf{X}_i)
 \]

 In practical classification tasks, perfect classification is unlikely. Hence, we need to balance interpretability with classification error.  Let $ \mathcal{E}_\mathbf{X} = \{(\mathbf{X}_i,y_i) | y_i \ne \mathcal{R}(\mathbf{X}_i) \} $   be  the set of samples in $ \mathbf{X} $ that are misclassified  by $ \mathcal{R} $. Therefore, we balance between classification-accuracy and rule-sparsity and optimize the following function.\footnote{In our formulation, it is  straightforward to add class-conditional weights  (e.g., to penalize  false-alarms more than mis-detects), and to allow instance weights (per sample).}

 \begin{equation}
  \label{interpretability_imli_eq:obj}
 \min\limits_{\mathcal{R}} \;  |\mathcal{E}_\mathbf{X}| +\lambda |\mathcal{R}|
 \end{equation}

>>>>

 \section{Problem Formulation}
 \label{interpretability_imli_sec:problem}
Given 
\begin{enumerate}
	\item a dataset $ \dataset = \{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n$ of $ n $ samples, where feature vector $ \mathbf{x}^{(i)} \in \{0, 1\}^m $ contains $ m $ features and class label $ y^{(i)} \in \{0,1\} $,
	\item a positive integer $ k \ge 1 $ denoting the number of clauses to be learned in the classification rule, and
	\item a regularization parameter $ \lambda \in \mathbb{R}^+ $,
\end{enumerate}
we learn a classifier $ \Rule $ represented as a $ k $-clause CNF formula that can separate samples of class $ 1 $ from class $ 0 $.

@@

 \[
 \min\limits_{\mathcal{R}} |\mathcal{R}|\text{ such that }\forall i, y^{(i)}=\mathcal{R}(\mathbf{x}^{(i)})
 \]

 In practical classification tasks, perfect classification is unlikely. Hence, we need to balance interpretability with classification error.  Let $ \mathcal{E}_\dataset = \{(\mathbf{x}^{(i)},y^{(i)}) | y^{(i)} \ne \mathcal{R}(\mathbf{x}^{(i)}) \} $   be  the set of samples in $ \dataset $ that are misclassified  by $ \mathcal{R} $. Therefore, we balance between classification-accuracy and rule-sparsity and optimize the following function.\footnote{In our formulation, it is  straightforward to add class-conditional weights  (e.g., to penalize  false-alarms more than mis-detects), and to allow instance weights (per sample).}

 \begin{equation}
  \label{interpretability_imli_eq:obj}
 \min\limits_{\mathcal{R}} \;  |\mathcal{E}_\dataset| +\lambda |\mathcal{R}|
 \end{equation}
